\section*{Contributions}

\paragraph{Overview.}
We address gradient aggregation for Reinforcement Learning with Verifiable Rewards (RLVR) under highly variable response lengths and realistic token dependencies. Our approach, \emph{Covariance-Aware Policy Optimization (CAPO)}, unifies prior length-based heuristics under a Bayesian perspective and introduces a simple, plug-in rule that yields unbiased, lower-variance updates without relying on token-independence assumptions.

\paragraph{Key contributions.}
\begin{itemize}
  \item \textbf{Unified analysis.} We place existing length-normalization rules (including $\Delta L$) on a coherent Bayesian footing via scale-invariant reasoning, clarifying the invariances and assumptions behind them.

  \item \textbf{Vector-valued optimality under common covariance shape.} We show that when per-trajectory covariances share a common ``shape'' up to a scalar factor, the inverse-variance rule remains optimal for a broad class of matrix risk functionals (Proposition~\ref{prop:common_shape_weights}), justifying scalar proxy modeling in high-dimensional policy gradients.
  \item \textbf{Robustness to variance misestimation.} We prove that plug-in inverse-variance weights degrade variance by at most a quadratic factor in the multiplicative error of the variance estimates (Proposition~\ref{prop:plugin_robust}), providing a formal stability guarantee under model misspecification.
  \item \textbf{CAPO aggregator.} We generalize beyond independence by estimating the \emph{induced variance law} $v(L)=\mathrm{Var}(g\mid L)$ under structured token dependence and using the corresponding inverse-variance (precision) weights, optionally combined with sublinear length dampening (\(\beta<1\)) to hedge variance-estimation noise.

  \item \textbf{Estimator-family view (submission-ready plug-ins).} We express CAPO as a modular GRPO-style reweighting rule that requires only a plug-in estimate of $v(L)$ (up to a common scale). This yields a coherent family of instantiations: \textbf{L-CAPO} (length-only/independence, including $\Delta L$ as the fixed-$\beta{=}1$ case), \textbf{LV-CAPO} (empirical Bayes with structured dependence), \textbf{CAPO-Q} (quadratic variance law motivated by compound symmetry / random-intercept covariance), and \textbf{CAPO-HAC} (lag-window/Newey--West long-run variance estimation).
  \item \textbf{Theory to practice.} We show that trajectory-level gradient variance grows approximately linearly with length under broad dependency classes, motivating CAPO’s form and its guidance for choosing $\beta<1$.
  \item \textbf{Simple implementation.} CAPO is a drop-in change at the level of group-based advantage reweighting: plug-ins produce per-trajectory precisions (or equivalently induced-variance estimates), and the main update normalizes these within each prompt group to obtain trajectory weights.
  \item \textbf{Diagnostics and robustness.} We provide practical checks (effective sample size, weight entropy, CV ratios, residual autocorrelation) to tune hyperparameters and detect covariance misfit.
  \item \textbf{Controlled synthetic covariance benchmark.} We provide a base-model-free simulator that generates length-heterogeneous rollout statistics under configurable within-trajectory covariance regimes and use it to compare GRPO, $\Delta L$, and CAPO variants directly at the level of advantage magnitudes and signal allocation across lengths.
  \item \textbf{Empirical protocol.} We specify a minimal, controlled evaluation suite (\S\ref{sec:results}) that compares CAPO against GRPO/DAPO/Dr.~GRPO and $\Delta L$ under matched tokens. We report both end accuracy and stability diagnostics, and include targeted ablations (sublinear exponent and covariance window size) to isolate covariance-aware effects.
\end{itemize}

\paragraph{Takeaway.}
CAPO offers a principled, covariance-aware alternative to length-only normalization: it preserves unbiasedness, reduces variance, and delivers stable, efficient RLVR training with minimal engineering effort.
